% +--------------------------------------------------------------------+
% | Acknowledgements Page (Optional)
% +--------------------------------------------------------------------+

\newpage
\vspace*{0.9cm}
\begin{center}
{\bf \Huge Acknowledgments}
\end{center}

\setlength{\baselineskip}{0.8cm}

%\pdfbookmark[0]{Acknowledgements}{PDF_Acknowledgements}

% +--------------------------------------------------------------------+
% | Enter text for your acknowledgements in the space below this box.
% |                                                                    
% +--------------------------------------------------------------------+
I would take some time to thank who has helped me at K-State.

First, I thank my parents, Xu Fajian and Lv Wei, who works as two distinguished mechanical engineers and pushed me to this great field.

I would like to thank to my advisor Professor Jeremy Roberts for providing me this great opportunity working on this project. He has served as my both undergraduate and graduate advisor in considerable patience since 2014.

I would like to thank Professor James Chen for leading me to numerical mathematics and encourage me to challenge myself.   

I would like to thank Professor Mingjun Wei and Hitesh Bindra for serving on my thesis committee and for answering any of my questions.

I would also like to thank two mentors and great friends, Dr. Mohammad Abdo and Richard Reed, for teaching me a great amount of expertise in nuclear engineering and data science.

I sincerely appreciate the help from my friends and colleagues at K-State - Cheikh, Dr. Wenkai Fu, Dr. Luis Wonnell, Khalid, Mohsen, Rabab, John, Ye.

For the funding for this effort, I acknowledge the Alan Levin Department of Mechanical and Nuclear Engineering department for sponsoring me in the past two years. 